---
title: "Structure of a gadget3 model"
output:
  html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Model Structure}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, message=FALSE, echo=FALSE}
library(gadget3)
library(magrittr)
```

The following describes the structure of a gadget3 model, from the bottom up.

## R formula, or the tilde operator

Crucial to gadget3 is the [R formula](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/formula.html),
created using the tilde operator (``~``). These get used in several places in R for
various things, but at their core the tilde operator stores the R code on the
left and right hand sides, as well as the environment it was created in, which
amounts to all variables defined at the time.

For example, let's declare a function that produces a formula, and make some:

```{r}
get_formula <- function (size) {
    # NB: The reason we make a function here is so we have an isolated environment
    # to make examples cleaner.
    cows <- size * 2
    pigs <- size * 4
    return(~cows + pigs)
}
f <- get_formula(8)
g <- get_formula(5)
```

``str()`` shows ``f`` contains the formula's code (not the result of cows + pigs),
and has an environment attached:

```{r}
str(f)
```

We can look into this environment, and see the values that got set for cows & pigs:

```{r}
str(as.list(environment(f)))
```

We can do similarly for ``g``, and see the results are different:

```{r}
str(as.list(environment(g)))
```

Note that:

* R hasn't at any point worked out ``cows + pigs``, the code was stored for
  later use.
* The environment (i.e. all variables defined at that point) is "remembered"

A g3 model at it's core is a list of formula objects that make up the model.
We can even use Gadget3 to compile our simple example above into an R function:

```{r}
g3_to_r(list(f))
```

...or a TMB template:

```{r}
g3_to_tmb(list(f))
```

Obviously this in itself isn't a very useful function, but you can see that the
environment for the formula has provided the initial values for the variables,
and the code itself has been put into the main loop. 

Also note that, despite appearances, we're not generically converting R into
C++. There's a subset of R that gadget3 understands and knows how to convert.
Using R libraries isn't possible, for example. To simplify scoping rules, we
assume variables are either global, and should have different names, or are
iterators in loops, in which case they will be local to that loop.

## Actions

In reality you will never be providing formulae to insert into models directly,
you'd be using the g3 action functions to generate these for you. All action
functions, prefixed with ``g3a_``, produce a list of formula objects---an
action in gadget3 parlance. These are where the gadget functionality is
implemented.

One of the simplest is ``g3a_time``, which produces code that will count
years/steps, and stop when the end of the time period is reached. For example:

```{r}
g3a_time(1990, 1999)
```

Like in the example above, the definitions are part of the formula's
environment, and if we compile it we see our years ending up in the code.

```{r}
g3_to_r(g3a_time(1990, 1999))
```

In this case our years have been hard-coded, but their definitions could
themselves be formula and the end result will be added to the model. For
example:

```{r}
g3_to_r(g3a_time(1990, ~start_year + 4 ))
```

TODO: This example rapidly runs out of steam since we don't substitute the
formula's environment into the output expression. Fix it or stop suggesting it.

## Stocks

Beyond ``g3a_time()``, pretty much any action will be describing changes to a
stock, possibly via. interacting with another stock (or fleet). To keep track
of this state, we use *g3_stock* objects. These describe several things:

* The dimensions one would use if making an array to store data on that stock.
For example if we want an array for number of individuals, what lengthgroups do
we use? How many ages do we store? How many (and which) areas do we have?
* What code do we need to iterate over the stock? Say I want to add 1
individual to each lengthgroup, how do I loop over all the other dimensions?
* If looping over one stock, how do I find corresponding entries in another
stock? For example, my fleet is only interested in prey that is in the same
area.

g3_stock objects can be created with either ``g3_stock()`` or ``g3_fleet()``,
the former will store lengthgroups, which fleets do not have.

Actions will store data about the stocks, for instance the current number of
individuals, in arrays called stock instances. We can see what sort of arrays a
stock will make by using ``gadget3:::stock_instance``:

```{r}
ling_imm <- g3_stock('ling_imm', seq(0, 50, 10))
gadget3:::stock_instance(ling_imm)
```

To add complexity to our model, we can use other ``g3s_`` functions, such as
``g3s_livesonareas()`` or ``g3s_age()``, which adds area or age dimensions to a
stock:

```{r}
ling_imm <- g3_stock('ling_imm', seq(0, 50, 10)) %>%
    g3s_age(3, 10)
gadget3:::stock_instance(ling_imm)
```

```{r}
ling_imm <- g3_stock('ling_imm', seq(0, 50, 10)) %>%
    g3s_livesonareas(c(1,2)) %>%
    g3s_age(3, 10)
gadget3:::stock_instance(ling_imm)[,,'age3']
```

When you use an action such as ``g3a_growmature()``, you provide g3_stock(s) to
act on, and formula objects to fill in gaps in the code. ``g3a_growmature()`` will
iterate over all areas/ages the stock has, and apply growth for each length
group it finds.

``g3a_growmature()`` itself doesn't care about areas or age, it
just does the same to each. However, the formula you supply can. ``age`` and
``area`` variables will be set with the current age/area, which you can use
when writing formula. For example:

```{r}
g3_to_r(g3a_growmature(
    ling_imm,
    growth_f = list(
        len = ~age * 10,
        wgt = ~area * 20),
    impl_f = g3a_grow_impl_bbinom(
        beta_f = ~g3_param("ling.bbin"),
        maxlengthgroupgrowth = 4),
    transition_f = ~TRUE))
```

...you can see our provided formula have been used to calculate
``ling_imm__growth_l`` and ``ling_imm__growth_w``, and ``age`` and ``area`` are
available to us thanks to the loops provided by the stock.

## Combining actions

Any useful model will have multiple actions, so the outputs from the ``g3a_``
functions need to be combined. To do this, you can pass a list of actions to
any of the ``g3_to_*`` functions, for example:

```{r}
ling_model <- g3_to_r(list(
    g3a_age(ling_imm),
    g3a_growmature(
        ling_imm,
        growth_f = list(
            len = ~age * 10,
            wgt = ~area * 20),
        impl_f = g3a_grow_impl_bbinom(
            beta_f = ~g3_param("ling.bbin"),
            maxlengthgroupgrowth = 4)),
    g3a_time(1990, 1999)))
```

A useful technique is to break down the actions into separate lists, e.g.

```{r}
ling_imm_actions <- list(
    g3a_age(ling_imm),
    g3a_growmature(
        ling_imm,
        growth_f = list(
            len = ~age * 10,
            wgt = ~area * 20),
        impl_f = g3a_grow_impl_bbinom(
            beta_f = ~g3_param("ling.bbin"),
            maxlengthgroupgrowth = 4)))
time_actions <- list(
    g3a_time(1990, 1999))

ling_model <- g3_to_r(c(ling_imm_actions, time_actions))
```

We do not need to worry about the order the actions are provided in, since they
will be re-ordered according to the [gadget2 order of
calculations](http://hafro.github.io/gadget2/userguide/chap-order.html).

This works as all formula within actions are named:

```{r}
sapply(ling_imm_actions, names)
sapply(time_actions, names)
```

...when building a model, formula are ordered by name, resulting in the desired
order of calculations. If needed, the ordering can be changed using the
``run_at`` parameter of any action.
