---
title: "Modelling maturity & sex with multiple stocks"
output:
  html_document:
    toc: true
    theme: null
vignette: >
  %\VignetteIndexEntry{Mutltiple stocks: maturity & sex}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{css, echo=FALSE}
/* https://bookdown.org/yihui/rmarkdown-cookbook/html-css.html */
.modelScript pre, .modelScript pre.sourceCode code {
  background-color: #f0f8ff !important;
}
```

```{r, message=FALSE, echo=FALSE}
library(unittest)
# Redirect ok() output to stderr
options(unittest.output = stderr())
if (nzchar(Sys.getenv('G3_TEST_TMB'))) options(gadget3.tmb.work_dir = gadget3:::vignette_base_dir('work_dir'))

library(gadget3)
set.seed(123)
```

This vignette walks through a script that will generate a gadget3 model,
explaining concepts along the way.

Code blocks that make up the gadget3 script are marked in blue, like this:

<div class="modelScript">
```{r, warning = FALSE, message = FALSE}
### Modelling maturity & sex with multiple stocks

```
</div>

When combined they will form a full model,
see [the appendix for the entire script](#appendix-full-model-script).

## Stocks & substocks

As mentioned before in `vignette('introduction-single-stock')`,
gadget3 stock objects do not have to correspond 1:1 with a species.

We can have multiple stock objects representing the same species in a different stage in their life-cycle,
most commonly `mature` and `immature` versions, `male` and `female` versions, or all 4.

The set up is much the same as before, but major differences will be highlighted.

Initial setup & time-keeping is identical:

<div class="modelScript">
```{r, warning = FALSE, message = FALSE}
library(gadget3)
library(dplyr)

actions <- list()
area_names <- g3_areas(c('IXa', 'IXb'))

# Create time definitions ####################

actions_time <- list(
  g3a_time(
    1979L, 2023L,
    step_lengths = c(3L, 3L, 3L, 3L)),
  NULL)

actions <- c(actions, actions_time)
```
</div>

### Stocks

We define 2 stocks instead of one, and a `list()` containing both for convenience:

<div class="modelScript">
```{r}
# Create stock definition for fish ####################
st_imm <- g3_stock(c(species = "fish", 'imm'), seq(5L, 25L, 5)) |>
  g3s_livesonareas(area_names["IXa"]) |>
  g3s_age(1L, 5L)

st_mat <- g3_stock(c(species = "fish", 'mat'), seq(5L, 25L, 5)) |>
  g3s_livesonareas(area_names["IXa"]) |>
  g3s_age(3L, 10L)
stocks = list(imm = st_imm, mat = st_mat)
```
</div>

Notice that:

* The name of the stock has 2 parts. This makes it possible to have parameters that refer to the species as a whole.
  In the model output the names will have been combined, e.g. ``"fish_imm"``.
* The age ranges are not identical, obviously mature stocks are older, and we adjust to suit.

### Stock actions

Stock actions now need to include interactions between immature & mature:

<div class="modelScript">
```{r}
actions_st_imm <- list(
  g3a_growmature(st_imm,
    g3a_grow_impl_bbinom(
      maxlengthgroupgrowth = 4L ),
    # Add maturation
    maturity_f = g3a_mature_continuous(),
    output_stocks = list(st_mat),
    transition_f = ~TRUE ),
  g3a_naturalmortality(st_imm),
  g3a_initialconditions_normalcv(st_imm),
  g3a_renewal_normalparam(st_imm),
  g3a_age(st_imm, output_stocks = list(st_mat)),
  NULL)

actions_st_mat <- list(
  g3a_growmature(st_mat,
    g3a_grow_impl_bbinom(
      maxlengthgroupgrowth = 4L )),
  g3a_naturalmortality(st_mat),
  g3a_initialconditions_normalcv(st_mat),
  g3a_age(st_mat),
  NULL)

actions_likelihood_st <- list(
  g3l_understocking(stocks, nll_breakdown = TRUE),
  NULL)

actions <- c(actions, actions_st_imm, actions_st_mat, actions_likelihood_st)
```
</div>

`actions_st_imm` and `actions_st_imm` are largely similar to our `actions_fish` from the previous model, but:

* We have added a `maturity_f` to `g3a_growmature()` to move individuals to the mature stock.
  The rate of maturity is coupled to growth, which is why `g3a_growmature()` does both at the same time.
* Immature `g3a_age()` can also move individuals to the mature stock.
  This will happen if an immature fish ages beyond the final age bin (5 in our case).
  At that point it matures "by default".
* Mature has no `g3a_renewal_normalparam()`, as there is no recruitment directly into the mature stock.

### Fleet actions

There is very little difference defining a fleet for a multiple stock model vs. single stocks.

To define a fleet, we need to introduce historical data into the model.
In our case we will generate random data to use later:

<div class="modelScript">
```{r}
# Fleet data for f_surv #################################

# Landings data: For each year/step/area
expand.grid(year = 1990:1994, step = 2, area = 'IXa') |>
    # Generate a random total landings by weight
    mutate(weight = rnorm(n(), mean = 1000, sd = 100)) |>
    # Assign result to landings_f_surv
    identity() -> landings_f_surv

# Length distribution data: Generate 100 random samples in each year/step/area
expand.grid(year = 1990:1994, step = 2, area = 'IXa', length = rep(NA, 100)) |>
  # Generate random lengths for these samples
  mutate(length = rnorm(n(), mean = 50, sd = 20)) |>
  # Save unagggregated data into ldist_f_surv.raw
  identity() -> ldist_f_surv.raw

# Aggregate .raw data
ldist_f_surv.raw |>
  # Group into length bins
  group_by(
      year = year,
      step = step,
      length = cut(length, breaks = c(seq(0, 80, 20), Inf), right = FALSE) ) |>
  # Report count in each length bin
  summarise(number = n(), .groups = 'keep') |>
  # Save into ldist_f_surv
  identity() -> ldist_f_surv

# Assume 5 * 5 samples in each year/step/area
expand.grid(year = 1990:1994, step = 2, area = 'IXa', age = rep(NA, 5), length = rep(NA, 5)) |>
  # Generate random lengths/ages for these samples
  mutate(length = rnorm(n(), mean = 50, sd = 20)) |>
  # Generate random whole numbers for age
  mutate(age = floor(runif(n(), min = 1, max = 5))) |>
  # Group into length/age bins
  group_by(
      year = year,
      step = step,
      age = age,
      length = cut(length, breaks = c(seq(0, 80, 20), Inf), right = FALSE) ) |>
  # Report count in each length bin
  summarise(number = n(), .groups = 'keep') ->
  aldist_f_surv
```
</div>

For more information on how this works, see `vignette("incorporating-observation-data")`.

Our fleet is defined with the same set of actions as the single-species model:

<div class="modelScript">
```{r}
# Create fleet definition for f_surv ####################
f_surv <- g3_fleet("f_surv") |> g3s_livesonareas(area_names["IXa"])

actions_f_surv <- list(
  g3a_predate_fleet(
    f_surv,
    stocks,
    suitabilities = g3_suitability_exponentiall50(by_stock = 'species'),
    catchability_f = g3a_predate_catchability_totalfleet(
      g3_timeareadata("landings_f_surv", landings_f_surv, "weight", areas = area_names))),
  NULL)
actions_likelihood_f_surv <- list(
  g3l_catchdistribution(
    "ldist_f_surv",
    obs_data = ldist_f_surv,
    fleets = list(f_surv),
    stocks = stocks,
    function_f = g3l_distribution_sumofsquares(),
    area_group = area_names,
    report = TRUE,
    nll_breakdown = TRUE),
  g3l_catchdistribution(
    "aldist_f_surv",
    obs_data = aldist_f_surv,
    fleets = list(f_surv),
    stocks = stocks,
    function_f = g3l_distribution_sumofsquares(),
    area_group = area_names,
    report = TRUE,
    nll_breakdown = TRUE),
  NULL)

actions <- c(actions, actions_f_surv, actions_likelihood_f_surv)
```
</div>

There are 2 differences to before:

* All actions use our list `stocks`, not an individual stock.
  Without additional changes, the 2 are treated as a single combined stock.
* We set `g3_suitability_exponentiall50(by_stock = 'species')`,
  instructing it to have a single `alpha` & `l50` parameter for both our stocks,
  as they have the same species name.

The `by_stock` parameter is passed through to `g3_parameterized()`.
We can see the result of setting this in the parameter template.
If ``by_stock = TRUE`` (the default) then we get parameters for both ``fish_imm.f_surv.l50`` & ``fish_mat.f_surv.l50``:

```{r}
simple_model <- g3_to_r(list(g3a_time(1990, 1994), g3a_predate_fleet(
    f_surv,
    stocks,
    suitabilities = g3_suitability_exponentiall50(by_stock = TRUE),
    catchability_f = g3a_predate_catchability_totalfleet(1) )))
names(attr(simple_model, "parameter_template"))
```

```{r, message=FALSE, echo=FALSE}
ok(ut_cmp_identical(sort(names(attr(simple_model, "parameter_template")), method = "radix"), c(
    "fish_imm.f_surv.alpha",
    "fish_imm.f_surv.l50",
    "fish_mat.f_surv.alpha",
    "fish_mat.f_surv.l50",
    "project_years", "retro_years",
    NULL)), "Params for simple_model, by_stock = TRUE")
```

If ``by_stock = 'species'``, then there is a single, shared ``fish.f_surv.l50`` parameter:

```{r}
simple_model <- g3_to_r(list(g3a_time(1990, 1994), g3a_predate_fleet(
    f_surv,
    stocks,
    suitabilities = g3_suitability_exponentiall50(by_stock = 'species'),
    catchability_f = g3a_predate_catchability_totalfleet(1) )))
names(attr(simple_model, "parameter_template"))
```

```{r, message=FALSE, echo=FALSE}
ok(ut_cmp_identical(sort(names(attr(simple_model, "parameter_template")), method = "radix"), c(
    "fish.f_surv.alpha",
    "fish.f_surv.l50",
    "project_years", "retro_years",
    NULL)), "Params for simple_model, by_stock = 'species'")
```

The ``by_stock`` parameter is just a convenient shortcut to change the default settings,
if we specify ``g3_parameterized()`` ourselves we can change the parameterization in other ways,
for example ``by_year = TRUE`` gives us per-year ``l50``:

```{r}
simple_model <- g3_to_r(list(g3a_time(1990, 1994), g3a_predate_fleet(
    f_surv,
    stocks,
    suitabilities = g3_suitability_exponentiall50(
        l50 = g3_parameterized("l50", by_stock = 'species', by_predator = TRUE, by_year = TRUE)),
    catchability_f = g3a_predate_catchability_totalfleet(1) )))
names(attr(simple_model, "parameter_template"))
```

```{r, message=FALSE, echo=FALSE}
ok(ut_cmp_identical(sort(names(attr(simple_model, "parameter_template")), method = "radix"), c(
    "fish.f_surv.l50.1990",
    "fish.f_surv.l50.1991",
    "fish.f_surv.l50.1992",
    "fish.f_surv.l50.1993",
    "fish.f_surv.l50.1994",
    "fish_imm.f_surv.alpha",
    "fish_mat.f_surv.alpha",
    "project_years", "retro_years",
    NULL)), "Params for simple_model, by_year = TRUE")
```

See ``?vignette('model-customisation')`` for more.

If we had data on the distribution of mature vs. immature, our observation data could contain a *stock*
column with ``fish_imm`` or ``fish_mat``. See `vignette("incorporating-observation-data")`.

Again, further fleets can be added by repeating the code above.

### Survey indices

Survey indices should be handed the full `stocks` list, instead of a single stock,
but are otherwise the same as before:

<div class="modelScript">
```{r}
# Create abundance index for si_cpue ########################

# Generate random data
expand.grid(year = 1990:1994, step = 3, area = 'IXa') |>
    # Fill in a weight column with total biomass for the year/step/area combination
    mutate(weight = runif(n(), min = 10000, max = 100000)) ->
    dist_si_cpue

actions_likelihood_si_cpue <- list(

  g3l_abundancedistribution(
    "dist_si_cpue",
    dist_si_cpue,

    stocks = stocks,
    function_f = g3l_distribution_surveyindices_log(alpha = NULL, beta = 1),
    area_group = area_names,
    report = TRUE,
    nll_breakdown = TRUE),
  NULL)

actions <- c(actions, actions_likelihood_si_cpue)
```
</div>

### Creating model functions and Parameterization

At this point, we are ready to convert our model into code:

<div class="modelScript">
```{r}
# Create model objective function ####################

model_code <- g3_to_tmb(c(actions, list(
    g3a_report_detail(actions),
    g3l_bounds_penalty(actions) )))
```
</div>

Now we should be configuring parameters based on the template.

Thanks to using wildcards in the `g3_init_val()` calls,
a lot of the parameter settings will work regardless of the model being single- or multi-stock,
so we don't need to change the initial values from the previous model:

<div class="modelScript">
```{r}
# Guess l50 / linf based on stock sizes
estimate_l50 <- g3_stock_def(st_imm, "midlen")[[length(g3_stock_def(st_imm, "midlen")) / 2]]
estimate_linf <- max(g3_stock_def(st_imm, "midlen"))
estimate_t0 <- g3_stock_def(st_imm, "minage") - 0.8

attr(model_code, "parameter_template") |>
  g3_init_val("*.rec|init.scalar", 10, lower = 0.001, upper = 200) |>
  g3_init_val("*.init.#", 10, lower = 0.001, upper = 200) |>
  g3_init_val("*.rec.#", 100, lower = 1e-6, upper = 1000) |>
  g3_init_val("*.rec.sd", 5, lower = 4, upper = 20) |>
  g3_init_val("*.M.#", 0.15, lower = 0.001, upper = 1) |>
  g3_init_val("init.F", 0.5, lower = 0.1, upper = 1) |>
  g3_init_val("*.Linf", estimate_linf, spread = 0.2) |>
  g3_init_val("*.K", 0.3, lower = 0.04, upper = 1.2) |>
  g3_init_val("*.t0", estimate_t0, spread = 2) |>
  g3_init_val("*.walpha", 0.01, optimise = FALSE) |>
  g3_init_val("*.wbeta", 3, optimise = FALSE) |>
  g3_init_val("*.*.alpha", 0.07, lower = 0.01, upper = 0.2) |>
  g3_init_val("*.*.l50", estimate_l50, spread = 0.25) |>
  g3_init_val("*.bbin", 100, lower = 1e-05, upper = 1000) |>
  identity() -> params.in
```
</div>

Finally we are ready for optimisation runs.
``g3_tmb_adfun()`` is a wrapper around ``TMB::MakeADFun()`` and ``TMB::compile``, producing a TMB *objective function*.
``gadgetutils::g3_iterative()`` then optimises based on iterative reweighting

<div class="modelScript">
```{r, eval=nzchar(Sys.getenv('G3_TEST_TMB'))}
# Optimise model ################################
obj.fn <- g3_tmb_adfun(model_code, params.in)

params.out <- gadgetutils::g3_iterative(getwd(),
    wgts = "WGTS",
    model = model_code,
    params.in = params.in,
    grouping = list(
        fleet = c("ldist_f_surv", "aldist_f_surv"),
        abund = c("dist_si_cpue")),
    method = "BFGS",
    control = list(maxit = 1000, reltol = 1e-10),
    cv_floor = 0.05)
```
</div>

Once this has finished, we can view the output using ``gadgetplots::gadget_plots()``.

<div class="modelScript">
```{r, eval=nzchar(Sys.getenv('G3_TEST_TMB'))}
# Generate detailed report ######################
fit <- gadgetutils::g3_fit(model_code, params.out)
gadgetplots::gadget_plots(fit, "figs", file_type = "html")
```
</div>

Once finished, you can view the output in your web browser:

<div class="modelScript">
```{r, eval=FALSE}
utils::browseURL("figs/model_output_figures.html")
```
</div>

## Appendix: Full model script

For convenience, here is all the sections of the model script above joined together:

```{js, echo=FALSE}
document.write(
    '<div class="modelScript"><div class="sourceCode hasCopyButton"><pre class="downlit sourceCode r">' +
    Array.from(document.querySelectorAll('.modelScript pre')).map((x) => x.innerHTML).join("\n\n") +
    '</pre></div></div>');
```

```{r, echo=FALSE, eval=nzchar(Sys.getenv('G3_TEST_TMB'))}
gadget3:::vignette_test_output("multiple-substocks", model_code, params.out)
```
