---
# Source vignette in R session:
#   Sys.setenv(G3_TEST_TMB=1)
#   knitr::purl('gadget3/vignettes/random-effects.Rmd', output="/tmp/vign.R") ; source("/tmp/vign.R", echo = T)
title: "Spawning & Random effects"
output:
  html_document:
    toc: true
    theme: null
vignette: >
  %\VignetteIndexEntry{Random effects: Spawning}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{css, echo=FALSE}
.hide {
  display: none!important;
}

/* https://bookdown.org/yihui/rmarkdown-cookbook/html-css.html */
.modelScript pre, .modelScript pre.sourceCode code {
  background-color: #f0f8ff !important;
}
```

```{r, message=FALSE, echo=FALSE}
library(unittest)
# Redirect ok() output to stderr
options(unittest.output = stderr())

library(gadget3)
set.seed(123)
```

<div class="alert alert-warning">
<strong>NB:</strong> This vignette is a work in progress
</div>

This vignette walks through a script that will generate a gadget3 model,
explaining concepts along the way.

Code blocks that make up the gadget3 script are marked in blue, like this:

<div class="modelScript">
```{r, warning = FALSE, message = FALSE}
### Spawning & Random effects

```
</div>

We only show the blocks that are relevant to implementing random effects,
see [the appendix for the entire script](#appendix-full-model-script).

For an explanation of other parts of the script, see previous vignettes.

<div class="modelScript hide">
```{r, warning = FALSE, message = FALSE}
library(gadget3)
library(dplyr)

actions <- list()
area_names <- g3_areas(c('IXa', 'IXb'))

# Create time definitions ####################

actions_time <- list(
  g3a_time(
    1980, 2000,
    step_lengths = c(3L, 3L, 3L, 3L)),
  NULL)

actions <- c(actions, actions_time)
```
</div>

<div class="modelScript hide">
```{r}
# Create stock definition for fish ####################
st_imm <- g3_stock(c(species = "fish", 'imm'), seq(5L, 25L, 5)) |>
  g3s_livesonareas(area_names["IXa"]) |>
  g3s_age(1L, 5L)

st_mat <- g3_stock(c(species = "fish", 'mat'), seq(5L, 25L, 5)) |>
  g3s_livesonareas(area_names["IXa"]) |>
  g3s_age(3L, 10L)
stocks = list(imm = st_imm, mat = st_mat)
```
</div>

## Spawning

In `vignette("introduction-single-stock")` we introduced `g3a_renewal_normalparam()`
as a means of adding recruitment into the model.

An alternative action is ``g3a_spawn()``.
The main difference here is the existing state of the stock is used to inform the rate of recruitment.

To improve our model fit, we shall define the rate of spawning using a random walk.

Firstly, we define our immature stock.
There's no real difference here, apart from ``g3a_renewal_normalparam()`` no longer being present:

<div class="modelScript">
```{r}
actions_st_imm <- list(
  g3a_growmature(st_imm,
    g3a_grow_impl_bbinom(
      maxlengthgroupgrowth = 4L ),
    # Add maturation
    maturity_f = g3a_mature_continuous(),
    output_stocks = list(st_mat),
    transition_f = ~TRUE ),
  g3a_naturalmortality(st_imm),
  g3a_initialconditions_normalcv(st_imm),
  # NB: g3a_renewal_normalparam() no longer here
  g3a_age(st_imm, output_stocks = list(st_mat)),
  NULL)
actions <- c(actions, actions_st_imm)
```
</div>

Next, we need to define the set of parameters that will constitute our random walk:

<div class="modelScript">
```{r}
# Configure our random walk for mu
st_spawn_mu <- g3_parameterized('spawn_mu', by_year = TRUE, random = TRUE)
```
</div>

As we saw in `vignette('multiple-substocks')`, ``by_year`` will give us a series of parameters for ``spawn_mu``, one per year.
``random = TRUE`` will ensure that these parameters have TRUE in the *random* column in the parameter template.

We store this as a separate variable as we need to use it in 2 places.
First, when we configure our spawning action:

<div class="modelScript">
```{r}
actions_st_mat <- list(
  g3a_growmature(st_mat,
    g3a_grow_impl_bbinom(
      maxlengthgroupgrowth = 4L )),
  g3a_naturalmortality(st_mat),
  g3a_initialconditions_normalcv(st_mat),
  g3a_spawn(
      st_mat,
      recruitment_f = g3a_spawn_recruitment_simplessb(mu = st_spawn_mu),
      #recruitment_f = g3a_spawn_recruitment_ricker(
      #    mu = st_spawn_mu,
      #    lambda = g3_parameterized("spawn_lambda", by_stock = TRUE) ),
      #    # TODO: Want seasonal spawning, so have to insert by_step somewhere.
      proportion_f = g3_suitability_exponentiall50(),
      output_stocks = list(st_imm) ),
  g3a_age(st_mat),
  NULL)
actions <- c(actions, actions_st_mat)
```
</div>

Mature stocks spawn, their output going into the immature stock.
We use suitability functions to define `proportion_f`, the proportion of stock ready to spawn,
``by_step`` meaning this proportion is seasonal.
Our random walk is used as the rate of recruitment, *mu*.

Next we use a likelihood action to constrain the values in the random walk during optimisation:

<div class="modelScript">
```{r}
actions_likelihood_st <- list(
  g3l_understocking(stocks, nll_breakdown = TRUE),
  g3l_random_walk('rwalk_st_spawn_mu',
      param_f = st_spawn_mu,
      sigma_f = g3_parameterized('spawn_mu.sigma', value = 5, optimise = FALSE),
      log_f = FALSE ),
  g3l_random_dnorm('rwalk_st_spawn_mu',
      param_f = st_spawn_mu,
      mean_f = g3_parameterized('spawn_mu.mean', value = 78, optimise = FALSE),
      sigma_f = g3_parameterized('spawn_mu.sigma', value = 1, optimise = FALSE),
      log_f = FALSE ),
  NULL)
actions <- c(actions, actions_likelihood_st)
```
</div>

For random effects to be useful, they have to be constrained to sensible bounds.
There are 2 likelihood actions that help with this, ``g3l_random_dnorm()`` & ``g3l_random_walk()``.
The former penalises random values that stray outside given normal distribution,
the latter constrains along a random walk; the current year's value should not vary significantly from the previous.

<div class="modelScript hide">
```{r}
# Fleet data for f_surv #################################

# Landings data: For each year/step/area
expand.grid(year = 1980:2000, step = 2, area = 'IXa') |>
    # Generate a random total landings by weight
    mutate(weight = rnorm(n(), mean = 1000, sd = 100)) |>
    # Assign result to landings_f_surv
    identity() -> landings_f_surv

# Length distribution data: Assume 100 samples in each year/step/area
expand.grid(year = 1980:2000, step = 2, area = 'IXa', length = rep(NA, 100)) |>
  # Generate random lengths for these samples
  mutate(length = rnorm(n(), mean = 50, sd = 20)) |>
  # Save our unaggregated data, carry on
  assign("ldist_f_surv.raw", value = _) |>
  # Group into length bins
  group_by(
      year = year,
      step = step,
      length = cut(length, breaks = c(seq(0, 80, 20), Inf), right = FALSE) ) |>
  # Report count in each length bin
  summarise(number = n(), .groups = 'keep') |>
  # Save into ldist_f_surv
  identity() -> ldist_f_surv

# Assume 5 * 5 samples in each year/step/area
expand.grid(year = 1980:2000, step = 2, area = 'IXa', age = rep(NA, 5), length = rep(NA, 5)) |>
  # Generate random lengths/ages for these samples
  mutate(length = rnorm(n(), mean = 50, sd = 20)) |>
  # Generate random whole numbers for age
  mutate(age = floor(runif(n(), min = 1, max = 5))) |>
  # Group into length/age bins
  group_by(
      year = year,
      step = step,
      age = age,
      length = cut(length, breaks = c(seq(0, 80, 20), Inf), right = FALSE) ) |>
  # Report count in each length bin
  summarise(number = n(), .groups = 'keep') ->
  aldist_f_surv
```
</div>

<div class="modelScript hide">
```{r}
# Create fleet definition for f_surv ####################
f_surv <- g3_fleet("f_surv") |> g3s_livesonareas(area_names["IXa"])

actions_f_surv <- list(
  g3a_predate_fleet(
    f_surv,
    stocks,
    suitabilities = g3_suitability_exponentiall50(by_stock = 'species'),
    catchability_f = g3a_predate_catchability_totalfleet(
      g3_timeareadata("landings_f_surv", landings_f_surv, "weight", areas = area_names))),
  NULL)
actions_likelihood_f_surv <- list(
  g3l_catchdistribution(
    "ldist_f_surv",
    obs_data = ldist_f_surv,
    fleets = list(f_surv),
    stocks = stocks,
    function_f = g3l_distribution_sumofsquares(),
    area_group = area_names,
    report = TRUE,
    nll_breakdown = TRUE),
  g3l_catchdistribution(
    "aldist_f_surv",
    obs_data = aldist_f_surv,
    fleets = list(f_surv),
    stocks = stocks,
    function_f = g3l_distribution_sumofsquares(),
    area_group = area_names,
    report = TRUE,
    nll_breakdown = TRUE),
  NULL)

actions <- c(actions, actions_f_surv, actions_likelihood_f_surv)
```
</div>

<div class="modelScript hide">
```{r}
# Create abundance index for si_cpue ########################

# Generate random data
expand.grid(year = 1980:2000, step = 3, area = 'IXa') |>
    # Fill in a weight column with total biomass for the year/step/area combination
    mutate(weight = runif(n(), min = 10000, max = 100000)) ->
    dist_si_cpue

actions_likelihood_si_cpue <- list(

  g3l_abundancedistribution(
    "dist_si_cpue",
    dist_si_cpue,

    stocks = stocks,
    function_f = g3l_distribution_surveyindices_log(alpha = NULL, beta = 1),
    area_group = area_names,
    report = TRUE,
    nll_breakdown = TRUE),
  NULL)

actions <- c(actions, actions_likelihood_si_cpue)
```
</div>

### Creating model functions and Parameterization

At this point, we are ready to convert our model into code:

<div class="modelScript">
```{r}
# Create model objective function ####################

model_code <- g3_to_tmb(actions)
```
</div>

<div class="modelScript">
```{r}
# Guess l50 / linf based on stock sizes
estimate_l50 <- g3_stock_def(st_imm, "midlen")[[length(g3_stock_def(st_imm, "midlen")) / 2]]
estimate_linf <- max(g3_stock_def(st_imm, "midlen"))
estimate_t0 <- g3_stock_def(st_imm, "minage") - 0.8

attr(model_code, "parameter_template") |>
  g3_init_val("*.rec|init.scalar", 10, lower = 0.001, upper = 200) |>
  g3_init_val("*.init.#", 10, lower = 0.001, upper = 200) |>
  g3_init_val("*.M.#", 0.15, lower = 0.001, upper = 1) |>
  g3_init_val("init.F", 0.5, lower = 0.1, upper = 1) |>
  g3_init_val("*.Linf", estimate_linf, spread = 0.2) |>
  g3_init_val("*.K", 0.3, lower = 0.04, upper = 1.2) |>
  g3_init_val("*.t0", estimate_t0, spread = 2) |>
  g3_init_val("*.walpha", 0.01, optimise = FALSE) |>
  g3_init_val("*.wbeta", 3, optimise = FALSE) |>
  g3_init_val("*.*.alpha", 0.07, lower = 0.01, upper = 0.2) |>
  g3_init_val("*.*.l50", estimate_l50, spread = 0.25) |>
  g3_init_val("*.bbin", 100, lower = 1e-05, upper = 1000) |>

  g3_init_val("spawn_mu.#", value = 78) |>
  #g3_init_val("spawn_mu.1980", value = 78, random = FALSE) |>
  #g3_init_val("spawn_lambda", value = 1e-6, optimise = TRUE) |>

  identity() -> params.in

# Uncomment this to temporarily disable random effects
#params.in[params.in$random, 'optimise'] <- TRUE
#params.in[params.in$random, 'random'] <- FALSE
```
</div>

Finally we are ready for optimisation runs.
``g3_tmb_adfun()`` is a wrapper around ``TMB::MakeADFun()`` and ``TMB::compile``, producing a TMB *objective function*.
``gadgetutils::g3_iterative()`` then optimises based on iterative reweighting

<div class="modelScript">
```{r, eval=nzchar(Sys.getenv('G3_TEST_TMB'))}
# Optimise model ################################
# Apply bounds in code - the other option would be using control = list(lower = g3_tmb_lower(params.in), ...)
bounded_code <- g3_to_tmb(c(actions, list(
    g3l_bounds_penalty(params.in) )))
obj.fn <- g3_tmb_adfun(bounded_code, params.in, inner.control = list(trace = 3, maxit = 100))

#obj.fn$env$tracepar <- TRUE
#obj.fn$env$tracemgc <- TRUE
out <- optim(par = obj.fn$par, fn = obj.fn$fn, gr = obj.fn$gr, method = 'BFGS', control = list(
    maxit = 1000,
    trace = 1,
    reltol = .Machine$double.eps^2 ))
params.out <- g3_tmb_relist(params.in, out$par)
fit <- gadgetutils::g3_fit(model_code, params.out)
```
</div>

## Appendix: Full model script

For convenience, here is all the sections of the model script above joined together:

```{js, echo=FALSE}
document.write(
    '<div class="modelScript"><div class="sourceCode hasCopyButton"><pre class="downlit sourceCode r">' +
    Array.from(document.querySelectorAll('.modelScript pre')).map((x) => x.innerHTML).join("\n\n") +
    '</pre></div></div>');
```
