---
# Source vignette in R session:
#   Sys.setenv(G3_TEST_TMB=1)
#   knitr::purl('gadget3/vignettes/random-effect-recruitment.Rmd', output="/tmp/vign.R") ; source("/tmp/vign.R", echo = T)
title: "Random effects & projections"
output:
  html_document:
    toc: true
    theme: null
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, message=FALSE, echo=FALSE}
library(unittest)
# Redirect ok() output to stderr
options(unittest.output = stderr())

library(gadget3)

set.seed(123)
```

Random effects in TMB run an inner optimisation problem for each run of the model being optimised.
To include random effects in a gadget3 model, you use the ``g3_param_project()`` series of functions.

``g3_param_project()`` works very much like ``g3_parameterized(by_year = TRUE)``, introducing a time-vector of parameters
into a model.
However, it also adds a likelihood component describing a relationship between each element of the time-vector,
which is then used to pick values for projections.

For instance using  ``g3_param_project("M", g3_param_project_dnorm())`` as the definition of "M" in
``g3a_naturalmortality()`` will result in the following parameters:

```{r, natural-mortality-example}
st <- g3_stock(c(species = "fish", "imm"), 1:10 * 10)
actions <- list(
    g3a_time(2000, 2003, c(6,6)),
    g3a_naturalmortality(st, g3a_naturalmortality_exp(
        g3_param_project("M", g3_param_project_dnorm()) )))
model_cpp <- g3_to_tmb(actions)
attr(model_cpp, "parameter_template")[,c("switch", "random", "source")]
```

* ``M.(year).(step)``: The time-vector of M values to use in historical timesteps
* ``M.proj.dnorm.mean`` / ``M.proj.dnorm.stddev``: The parameters of the normal distribution we expect ``M`` to meet. These will...
  * Add the time-vector's deviance from the ideal distribution to the model likelihood, so optimised solutions should fit the distribution.
  * Be used to select future values of ``M`` when projecting
* ``proj_dnorm_M_weight``: The weighting of the likelihood component used above

## Defining recruitment as a random effect

For a more complete example, we create a single-stock model, with a g3l_abundancedistribution() likelihood component enforcing constant abundance.
We define the renewal rate using ``g3_param_project()`` with ``g3_param_project_logar1()``:

```{r, model-setup}
stocks <- list(
    imm = g3_stock(c(species = "fish", "imm"), 1:10 * 10) |> g3s_age(1, 5) )

## Generate some abundnace data for ages 4 & 5, the stock should be stable, however the data is noisy
old.seed <- mget(c(".Random.seed"), envir = globalenv(), ifnotfound = list(NULL))[[1]]
set.seed(42)
dist_age4 <- expand.grid(year = 2000:2030, step = 1, age = 4, number = NA)
dist_age4$number <- rnorm(nrow(dist_age4), 1.75e7, 2e6)
dist_age5 <- expand.grid(year = 2000:2030, step = 1, age = 5, number = NA)
dist_age5$number <- rnorm(nrow(dist_age5), 3.75e7, 2e6)
set.seed(old.seed)

## Single stock model, using our generated data as an abundance index
actions <- list(
    g3a_time(2000, 2030, step_lengths = c(6,6)),
    g3a_age(stocks$imm),
    g3a_initialconditions_normalcv(stocks$imm),
    g3a_naturalmortality(stocks$imm),
    g3a_growmature(stocks$imm, g3a_grow_impl_bbinom(maxlengthgroupgrowth = 4L)),
    g3a_renewal_normalparam(
      stocks$imm,
      factor_f = g3_param_project(
        "rec",
        g3_param_project_logar1(),
        scale = "scalar",
        by_stock = stocks,
        by_step = FALSE ),
      run_step = 1 ),

    g3l_abundancedistribution(
        "dist_older",
        rbind(dist_age4, dist_age5),
        stocks = stocks,
        # NB: Fix alpha/beta, abundance should match given values absolutely
        function_f = g3l_distribution_surveyindices_log(alpha = 0, beta = 1),
        report = TRUE,
        nll_breakdown = TRUE),
    NULL )

full_actions <- c(actions, list(
    g3a_report_detail(actions),
    g3a_report_history(actions, "__num$|__wgt$", out_prefix="dend_"),  # NB: Late reporting
    g3l_bounds_penalty(actions),
    NULL ))
model_cpp <- g3_to_tmb(full_actions)
```

Next, we define some initial parameters to use.

Note in particular that we are optimising the parameters for the distribution governing the time-vector.
The aim here is roughly that
the outer optimisation will decide on values here that meet the overall model,
the inner optimistation will choose values that meet the distribution.

```{r,  model-parameterisation}
## Define baseline parameters / fixed-effects
attr(model_cpp, "parameter_template") |>
  g3_init_val("*.init.scalar", 10, optimise = FALSE) |>
  g3_init_val("*.init.#", 10, lower = 1e1, upper = 1e8) |>
  g3_init_val("*.M.#", 0.5, lower = 0.1, upper = 1) |>
  g3_init_val("init.F", 0.5, lower = 0.1, upper = 10) |>
  g3_init_val("*_imm.Linf", 144.645) |>
  g3_init_val("*.K", 0.3, lower = 0.04, upper = 1.2) |>
  # NB: We don't optimise here since the model defines no length/weight structure, only counts
  g3_init_val("*.t0", -0.8, optimise = FALSE) |>
  g3_init_val("*.walpha", 0.01, optimise = FALSE) |>
  g3_init_val("*.wbeta", 3, optimise = FALSE) |>
  g3_init_val("*.rec.scalar", 10, optimise = FALSE) |>
  g3_init_val("fish_imm.rec.proj.logar1.level", value = 2e3, lower = 1e1, upper = 5e5) |>
  g3_init_val("fish_imm.rec.proj.logar1.stddev", value = 1, lower = 1e-15, upper = 1e4) |>
  # NB: Relative weight is important, but g3_iterative isn't handling this for you yet
  g3_init_val("proj_logar1_fish_imm_weight.proj_logar1_rec_weight", 1e0) |>
  identity() -> params.in
```

To give something to compare against, optimise first with recruitment as a fixed-effect,
which should still work, as the score for fixed & random effects is combined:

```{r, optimisation-fixed, cache=FALSE}
params.in |>
  # NB: level = init value, to be internally consistent
  g3_init_val("*.rec.#", value = 2e3, lower = 1e1, upper = 1e6, random = FALSE) |>
  identity() -> params.fixed
obj.fix <- g3_tmb_adfun(model_cpp, params.fixed, silent = TRUE)
params.fixout <- gadgetutils::g3_optim(obj.fix, params.fixed, control = list(maxit = 2000, trace = 0))
stopifnot(attr(params.fixout, "summary")$convergence)
```

Next, optimise the same model but using random effects for recruitment:

```{r, optimisation-random, cache=FALSE}
params.in |>
    # NB: By setting random = TRUE, any previous bounds will be cleared
    g3_init_val("*.rec.#", value = 2e3, random = TRUE) |>
    identity() -> params.rnd
obj.rnd <- g3_tmb_adfun(model_cpp, params.rnd, inner.control = list(trace = 0, maxit = 1000, tol = 1e-3), silent = TRUE)
params.rndout <- gadgetutils::g3_optim(obj.rnd, params.rnd, trace = 0)
stopifnot(attr(params.rndout, "summary")$convergence)
```

Both models have ended up with similar solutions:

```{r, params-out}
print(cbind(params.fixout$value, params.rndout$value))
```

```{r, sdreport, eval=FALSE, echo=FALSE}
# TODO: TMB::sdreport only works with nlminb(), not g3_optim(), or optim(), but we're getting false convergence
obj.rndnlm <- g3_tmb_adfun(model_cpp, params.rnd, inner.control = list(trace = 0, maxit = 1000, tol = 1e-3))
nlmfit <- nlminb(obj.rndnlm$par, obj.rndnlm$fn, obj.rndnlm$gr, lower = g3_tmb_lower(params.rndout), upper = g3_tmb_upper(params.rnd),
    control = list(eval.max = 2000, iter.max = 2000, trace = 0))
summary(TMB::sdreport(obj.rndnlm))
```

## Projections

Given a parameterised model, projections can be performed by varying the ``project_years`` parameter.
Whilst you can use the objective function above for this, it's simpler to do so with a function designed for single runs,
produced either by ``g3_tmb_fn`` or ``g3_tmb_r``. We will use the latter here:

```{r, single-run-function}
fn <- g3_tmb_fn(model_cpp)
```

The resulting ``fn`` function then accepts parameter tables, which we can modify to project into the future.
Our model settles at the point we expect:

```{r, mean-abundance}
abund <- g3_array_agg(
    fn(params.rndout |> g3_init_val("project_years", 40))$dend_fish_imm__num,
    c('age', 'year'),
    age = c(4, 5),
    year = 2040:2060,
    step = 1)
print(signif(rowMeans(abund), 3))
```

```{r, message=FALSE, echo=FALSE}
ok(ut_cmp_equal(rowMeans(abund), c(age4 = 1.75e+07, age5 = 3.75e+07), tolerance = 2e-2), "Tolerance at equlibrium point")
```

Plot abundance for years 4 & 5, with 40 years projection, using level/stddev chosen by optimiser:

```{r, proj-plot-all-years}
r <- fn(params.rndout |> g3_init_val("project_years", 40))
g3_array_plot(t(g3_array_agg(r$dend_fish_imm__num, c('age', 'year'))))
```

Successive calls will choose different values for the projections.
Plot 20 iterations of projecting 40 years:

```{r, proj-plot-20-iters}
proj_iters <- do.call(cbind, lapply(1:20, function (i) {
    r <- fn(params.rndout |> g3_init_val("project_years", 40))

    # Generate report for age 4 abundance
    proj <- t(g3_array_agg(
        r$dend_fish_imm__num,
        c('age', 'year'),
        year = 2025:2060,  # NB: Skip the first years when the model is settling
        age = 4,
        step = 1 ))
    dimnames(proj)[[2]] <- paste0("proj ", i)
    return(proj)
}))
g3_array_plot(proj_iters)
```
